---
title: "Data607 - Week 2"
author: "Sherranette Tinapunan"
date: "February 11, 2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Homework 2

>Choose six recent popular movies.  Ask at least five people that you know (friends, family, classmates, imaginary friends) to rate each of these movie that they have seen on a scale of 1 to 5.  Take the results (observations) and store them in a SQL database.  Load the information into an R dataframe.

> This is by design a very open ended assignment.  A variety of reasonable approaches are acceptable.  You can (and should) blank out your SQL password if your solution requires it; otherwise, full credit requires that your code is "reproducible," with the assumption that I have the same database server and R software.


### Tutorial 

> Source: http://www.datacarpentry.org/R-ecology-lesson/05-r-and-databases.html

### Packages

* RMySQL --> package that allows us to connect to MySQL database directly from R
* dplyr --> allows us to query/retrieve data from database as needed
* dbplyr --> allows us to edit database directly

* install the dbplyr for RMySQL 
* Code: install.packages(c("dbplyr", "RMySQL"))


### Load libraries
```{r}
library(RMySQL)
library(dplyr)
library(dbplyr)
```

### Connect to MySQL database running locally

> I am connecting to a MySQL server that is local to my computer at home. 
The "data607_hw2" MySQL database has 2 tables: movies, and movie_ratings. 

```{r}
con <- DBI::dbConnect(RMySQL::MySQL(), dbname = "data607_hw2", user="root", password="P@55w0rd123!")

#get info on the database
src_dbi(con)
```


### Query tables and create data frame

```{r}
#movies table
movies <- tbl(con, sql("SELECT movie_id, movie_title, movie_year FROM movies"))

#ratings table
ratings <- tbl(con, sql("SELECT * FROM movie_ratings"))

#JOIN of both movies and ratings table
movies_ratings <- tbl(con, sql("SELECT a.rating_id, a.rating, b.movie_title, b.movie_year
             FROM movie_ratings a JOIN movies b ON a.movie_id = b.movie_id"))

movies_ratings_df <- as.data.frame(movies_ratings)

#data frame: 
dim(movies_ratings_df)
movies_ratings_df
```
